---
title: "Sample Size Calculator for Micro-Randomized Trials"
author: "Thabat Dahdoul, Eliot Wong-Toi"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample Size Calculator for Micro-Randomized Trials (MRTs)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
bibliography: vig_bib.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mrtbincalc)
```

# Introduction 
The **mrtbincalc** package provides a sample size calculator for
micro-randomized trials (MRT) with binary outcomes. MRTs are an experimental design
for optimizing mobile health interventions such as push notifications to increase physical activity.
The sample size calculator is developed in @cohn2021sample.
For a general overview of MRT, see @klasnja_microrandomized_2015 and @liao2016sample.

This calculator provides the sample size needed
for a specified power to test for a nonzero marginal causal excursion effect of the
mobile health intervention, and can also be used to find the power given a sample size. 

Here we briefly review the hypothesis test that the sample size is calculated for.
Let $(X_1, A_1, Y_2, X_2, A_2, Y_3, \ldots, X_m, A_m, Y_{m+1})$ denote the observed data
from a participant in the MRT.
$m$ denotes the total number of decision points.
For each decision point $t$, $X_t$ denotes the time-varying covariate,
$A_t$ denotes the treatment assignment,
and $Y_{t+1}$ denotes the binary proximal outcome.
We assume $A_t$ takes value in $\{0, 1\}$: For example, if the treatment is 
a push notification, then $A_t = 0$ means the push notification is not delivered at $t$,
and $A_t = 1$ means the push notification is delivered at $t$.
Let $I_t$ denote the availability condition at $t$: $I_t = 1$ if the participant is
available for treatment at $t$, and $I_t = 0$ if the participant is unavailable at $t$.

For primary analysis of an MRT, one is usually interested in testing for a marginal causal
excursion effect (MEE). Under standard causal assumptions, MEE at decision point $t$ can be expressed as
$$
\text{MEE}(t) = \log \frac{P(Y_{t+1} = 1 \mid A_t = 1, I_t = 1)}{P(Y_{t+1} = 1 \mid A_t = 0, I_t = 1)},
$$
which is akin to a relative risk. One is interested in testing the following null hypothesis
$$
H_0: \text{MEE}(t) = 0 \text{ for all } 1\leq t \leq m
$$
against the alternative hypothesis
$$
H_1: \text{MEE}(t) \neq 0 \text{ for some } 1\leq t \leq m.
$$

The sample size is calculated to ensure adaquate power to detect a particular target alternative
$$
H_1^{\text{target}}: \text{MEE}(t) = f(t)^T \beta \text{ for all } 1\leq t \leq m,
$$
where $f(t)$ is a user-specified $p$-dimensional vector-valued function of $t$ and $\beta \in \mathbb{R}^p$. For example, if the user conjectures that a likely alternative is a constant treatment effect, they can set $f(t) = 1$ and $\beta$ will be a scalar; if the user conjectures that the treatment effect would decrease over time, they can set $f(t) = (1, t)^T$ and $\beta$ will be a vector of length 2. The user will supply $f(t)$ and $\beta$, along with other inputs that will be detailed below, when using the calculator in this package.


# Quick Start
This section provides examples of how to use this package. It allows the user to
see the basic operations and what the outputs would look like. The functions in
this package rely on the some specific information from the MRT set up, such as
the number of decision time points over the course of the study, and the the
probability of assigning the treatment at each decision time point (the
randomization probability).

Since in an MRT treatment can only be provided when an individual is available,
we must account for this. The expected availability is the probability a person
is available to receive the intervention at a specific decision time. To use
this package, you need to define a time-varying pattern for the expected
availability.

Further pre-specified information that the user needs to provide is the success
probability null curve at each decision time point. It is defined as the
probability of the proximal outcome equal to 1 for available individuals who are
not assigned treatment. The user needs to provide the trend of success
probability of the null curve. For example, this could be either constant or
linear over decision points.


Finally, the proximal treatment effect at each decision time point is defined as
the mean difference in the proximal outcome between available people who are
assigned a treatment versus available people who are not assigned treatment. In
this work, we only consider the binary treatment. One needs to provide the trend
of proximal treatment effects.


The computation of the required sample size/power relies on a test statistic
that is distributed as a non-central $F$ with $p, n-q-p$ degrees of freedom
under the alternative hypothesis. For specifics about this see @pan2002small.
 

The two main operations of this package are to find the sample size required for
a specified power and to find the power for a specified sample size in the
context of an MRT.

## Data
The following data is included in the package and will be used to illustrate
some of the functions in this package:
 

`tau_t_1`: Vector of length 10 that holds the average availability at each time
point. In this example the availability remains constant across decision points.
```{r}
tau_t_1
```


`f_t_1`: A 10 by 2 matrix defining the alternative hypothesis. In this example
it is a log-linear trend. Each row corresponds to a decision point (in this
example we have 10).
```{r}
f_t_1
```


`g_t_1`: A 10 by 2 matrix defining the null hypothesis.In this example it is a
log-linear trend. As with `f_t_1`, each row corresponds to a decision point.
```{r}
g_t_1
```


`p_t_1`: A length 10 vector of randomization probabilities for each time point. 
This example has the randomization probability staying constant across decision 
points.
```{r}
p_t_1
```


`beta_1`: Vector that defines the alternative hypothesis.The matrix
multiplication of this vector with f_t_1 defines the MEE under the alternative
hypothesis.
```{r}
beta_1
```


`alpha_1`: A length of 2 vector that defines the null hypothesis.The matrix
multiplication of this vector with g_t_1 defines the MEE under the null
hypothesis.
```{r}
alpha_1
```

## Examples
### `calculate_mrt_bin_samplesize_f()`
Below we compute the required sample size for an MRT with the above data to
achieve $0.8$ power. The argument `gamma` is the type I error rate, `b` is the
type II error rate ($1-power$), and `exact` is a flag for if the function should
return the exact sample size our calculator computes (this may not be an
integer) or the ceiling of this number. By default `exact=FALSE`. We see that
the required sample size is $275$ individuals for our setup.
```{r}
calculate_mrt_bin_samplesize_f(avail_pattern=tau_t_1, 
                               f_t=f_t_1, g_t=g_t_1, 
                               beta=beta_1, alpha=alpha_1,
                               p_t=p_t_1, 
                               gamma=0.05, 
                               b=.2,
                               exact=FALSE)
```

### `calculate_mrt_bin_power_f()`
If the investigator would like to calculate power based on a fixed sample size
that is feasible in the study, with a specified significance level, then they
can use the `calculate_mrt_bin_power_f()` function. The first seven arguments
are the same as in `calculate_mrt_bin_samplesize_f()`. The final argument, `n`
is the number of individuals. Notice how the sample size (`n=275`) is the output
from the previous computation and the power is very close to $0.8$, there is a
slight discrepancy due to the rounding up of the sample size in the previous
example.
```{r}
calculate_mrt_bin_power_f(avail_pattern=tau_t_1, 
                          f_t=f_t_1, g_t=g_t_1, 
                          beta=beta_1, alpha=alpha_1, 
                          p_t=p_t_1, 
                          gamma=0.05, 
                          n=275)
```
### `power_vs_n_plot()`
Based on the fraction of power calculated above, the user can employ this
information to trade off/make the decision on power vs. sample size. This
decision can be costly, and so having a visualization of this relationship
between power and sample size can be beneficial.

Thus, the package provides a visualization of this relationship and so we
recommend the user to use the `power_vs_n_plot()` function. This example uses
the default range of sample sizes to plot over, but with the `min_n` and `max_n`
the user can choose what range they want to plot over.

```{r powe-plot, fig.align='center'}
power_vs_n_plot(avail_pattern=tau_t_1, 
                f_t=f_t_1, g_t=g_t_1, 
                beta=beta_1, alpha=alpha_1, 
                p_t=p_t_1, 
                gamma=0.05)
```

### `power_summary()`
Another way to look at this information that quantifies power in relation to
sample size, is by looking at a list of specific powers and the sample size
needed to achieve it. This package offers a `power_summary()` function that
considers sample sizes for power ranging from $0.6$ to $0.95$ by increments of
$0.05$ by default. Again, we see that a sample size of $275$ will achieve a
power of $0.8$ with these settings. If the user is interested in values not
included in this range they can input any sequence of power levels in the
`power_levels` argument.
```{r}
power_summary(tau_t_1, f_t_1, g_t_1,beta_1, alpha_1, p_t_1, 0.05)
```




# References







